---
title: "Visualization Journal"
subtitle: "CSC 477 / GCSC 577"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
jupyter: python3
---

## About

This site is my course notebook for CSC 477 / GCSC 577. It contains weekly notes, small experiments, and visualizations created throughout the semester.

## Week 1

### Notes

- 

### Visualization

```{r}
library(ggplot2)

ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  labs(
    title = "Horsepower vs. MPG",
    x = "Miles per gallon (mpg)",
    y = "Horsepower (hp)"
  )
```

### Quick reflection

- What stands out:
- What I want to explore next:

## Week 2

### Notes

- 

### Visualization

```{python}
import pandas as pd
import matplotlib.pyplot as plt

df = pd.DataFrame({
    "x": [1, 2, 3, 4, 5],
    "y": [2, 1, 3, 5, 4],
})

plt.figure()
plt.plot(df["x"], df["y"], marker="o")
plt.title("Simple Line Plot")
plt.xlabel("x")
plt.ylabel("y")
plt.show()
```

### Quick reflection

- What stands out:
- What I want to explore next:

## Week 3

### Notes

- 

### Visualization

```{r}
library(dplyr)
library(ggplot2)

summary_tbl <- mtcars %>%
  group_by(cyl) %>%
  summarize(
    avg_mpg = mean(mpg),
    n = n(),
    .groups = "drop"
  )

ggplot(summary_tbl, aes(x = factor(cyl), y = avg_mpg)) +
  geom_col() +
  labs(
    title = "Average MPG by Cylinder Count",
    x = "Cylinders",
    y = "Average MPG"
  )
```

### Quick reflection

- What stands out:
- What I want to explore next:
