---
title: "Week 06 — DataWeb: U.S. Exports (2010–2024)"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

<!--
==============================================================
FILE: week-06.qmd
PROJECT: Visualization Journal (CSC 477 / GCSC 577)
AUTHOR: Salmon Riaz
VERSION: 1.0.2
DATE: 2026-02-27
CHANGE (v1.0.2): Fixed R regex parsing error by replacing \d escapes with [0-9] (safe in R strings).
CHANGE (v1.0.1): Fixed R regex parsing error by replacing \s escapes with [[:space:]].
==============================================================

WHY THIS CHANGE?
- R treats "\s" inside normal strings as an escape sequence and errors during parsing.
- Using the POSIX whitespace class [[:space:]] avoids the issue and is portable.

NOTES
- DataWeb exports sometimes have “header rows” a few lines down; this file tries to locate headers
  defensively by searching for year-columns like 2010, 2011, etc.
- If your DataWeb download changes the sheet name away from "FAS Value", update the sheet= value
  below. The `excel_sheets()` calls help you confirm what’s in each workbook.
-->

## DataWeb exports dataset {#dataweb}

This section uses two exports datasets downloaded from the U.S. International Trade Commission (USITC) DataWeb site:

- **Total U.S. exports (all destinations)**, 2010–2024  
- **U.S. exports by destination country**, 2010–2024  

Both spreadsheets are stored in this repo under `data/` so the site can render the analysis reproducibly.

::: {.callout-note}
If your DataWeb download includes slightly different sheets/headers in the future, the cleaning steps below are written defensively (they search for the “2010…2024” year columns and build tidy data from them).
:::

## 1) Load packages

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

## 2) Read the Excel files (all sheets)

```{r}
# File paths (relative to project root)
f_total   <- "data/DataWeb-Query-Export_us_exports_total_2010_2024.xlsx"
f_country <- "data/DataWeb-Query-Export_us_exports_by_country_2010_2024.xlsx"

# Optional: list every sheet (helps satisfy the “look at all sheets” requirement)
excel_sheets(f_total)
excel_sheets(f_country)
```

## 3) Total exports (tidy to Year + Value)

```{r}
# The “FAS Value” sheet is the main data table in both downloads.
# DataWeb often places the real header row a couple rows down.
raw_total <- read_excel(f_total, sheet = "FAS Value", col_names = FALSE)

# Find the header row by looking for the first row that contains a year like 2010
header_row_total <- which(apply(raw_total, 1, function(r) any(str_detect(as.character(r), "^2010$"))))[1]
header_row_total
```

```{r}
# Build column names from that header row, then keep the data rows under it.
colnames_total <- str_trim(as.character(unlist(raw_total[header_row_total, ])))
colnames_total <- make.unique(colnames_total)
total_df <- raw_total[(header_row_total + 1):nrow(raw_total), ]
names(total_df) <- colnames_total

# IMPORTANT FIX:
# - Use POSIX whitespace class instead of "\s" to avoid R parse errors.
year_col_pattern <- "^[[:space:]]*[0-9]{4}[[:space:]]*$"

# Keep only the year columns and pivot longer
total_tidy <- total_df |>
  select(matches(year_col_pattern)) |>
  slice(1) |>
  pivot_longer(
    cols = everything(),
    names_to = "Year",
    values_to = "Exports"
  ) |>
  mutate(
    Year = as.integer(str_trim(Year)),
    Exports = as.numeric(Exports)
  ) |>
  arrange(Year)

glimpse(total_tidy)
```

## 4) Exports by country (tidy to Country + Year + Value)

```{r}
raw_country <- read_excel(f_country, sheet = "FAS Value", col_names = FALSE)

# Find the header row by looking for the first row that contains "Country"
header_row_country <- which(apply(raw_country, 1, function(r) any(str_detect(as.character(r), "^Country$"))))[1]
header_row_country
```

```{r}
colnames_country <- str_trim(as.character(unlist(raw_country[header_row_country, ])))
colnames_country <- make.unique(colnames_country)
country_df <- raw_country[(header_row_country + 1):nrow(raw_country), ]
names(country_df) <- colnames_country

# Reuse the same year column pattern
year_col_pattern <- "^[[:space:]]*[0-9]{4}[[:space:]]*$"

country_tidy <- country_df |>
  filter(!is.na(Country)) |>
  select(Country, matches(year_col_pattern)) |>
  pivot_longer(
    cols = -Country,
    names_to = "Year",
    values_to = "Exports"
  ) |>
  mutate(
    Country = str_trim(as.character(Country)),
    Year = as.integer(str_trim(Year)),
    Exports = as.numeric(Exports)
  ) |>
  filter(!is.na(Exports)) |>
  arrange(Country, Year)

glimpse(country_tidy)
```

## 5) Visualizations

### A. Total U.S. exports over time (line chart)

```{r}
ggplot(total_tidy, aes(x = Year, y = Exports)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  labs(
    title = "Total U.S. Exports (FAS Value), 2010–2024",
    subtitle = "Source: USITC DataWeb download",
    x = "Year",
    y = "Exports (FAS Value, reported units from DataWeb download)"
  ) +
  theme_minimal()
```

### B. Top 10 export destinations in 2024 (bar chart)

```{r}
top_2024 <- country_tidy |>
  filter(Year == 2024) |>
  group_by(Country) |>
  summarise(Exports = sum(Exports, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(Exports)) |>
  slice_head(n = 10)

top_2024
```

```{r}
ggplot(top_2024, aes(x = reorder(Country, Exports), y = Exports)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 U.S. Export Destinations (2024)",
    subtitle = "FAS Value from USITC DataWeb download",
    x = "",
    y = "Exports (FAS Value, reported units from DataWeb download)"
  ) +
  theme_minimal()
```

### C. Share of top destinations vs the world (2024)

```{r}
total_2024 <- total_tidy |>
  filter(Year == 2024) |>
  pull(Exports) |>
  as.numeric()

share_2024 <- top_2024 |>
  mutate(Share = Exports / total_2024) |>
  arrange(desc(Share))

share_2024
```

```{r}
ggplot(share_2024, aes(x = reorder(Country, Share), y = Share)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Share of Total U.S. Exports (2024): Top 10 Destinations",
    subtitle = "Shares computed using Total Exports (all destinations) from the companion DataWeb download",
    x = "",
    y = "Share of total exports"
  ) +
  theme_minimal()
```

## Quick interpretation (STAR: Scope + Action)

- **Scope (context):** This is annual U.S. export value (FAS Value) from DataWeb, covering 2010–2024.  
- **Action (what we did):** We pulled the total series and the country breakdown into tidy tables, then visualized trends over time and the top destinations in the most recent full year.
