---
title: "Week 07 — Multivariate Correlation Analysis (U.S. Exports Project)"
author: "Salmon Riaz"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This notebook applies **multivariate correlation** to a U.S. exports dataset (DataWeb export totals).
The goal is to understand how major export destinations **move together over time**, and to translate that
structure into a practical next step that connects exports to an **Economy & Labor** indicator.

## Objectives

1. Load and clean the exports dataset (robustly, even if Excel sheet layout varies).
2. Reshape the data into:
   - **Long format** (for time-series plotting)
   - **Wide numeric format** (for correlation analysis)
3. Create multivariate visuals:
   - Time-series trends for top export destinations
   - Correlation matrix + heatmap
   - `corrplot` (if installed)
   - Scatterplot matrix (`GGally::ggpairs`, if installed)
4. **Economy & Labor tie-in (implemented):**
   - Merge annual exports with an annual labor indicator
   - Run correlation + a simple regression to quantify the relationship

---

## 1) Setup

```{r}
library(tidyverse)
library(readxl)
library(stringr)
library(scales)

# Optional helpers (loaded only if installed)
has_pkg <- function(pkg) {
  requireNamespace(pkg, quietly = TRUE)
}
use_pkg <- function(pkg) {
  if (has_pkg(pkg)) {
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
    TRUE
  } else {
    FALSE
  }
}

# -----------------------------
# Helper: pick first path that exists
# -----------------------------
first_existing <- function(paths) {
  hit <- paths[file.exists(paths)]
  if (length(hit) == 0) NA_character_ else hit[1]
}

# -----------------------------
# Helper: identify columns that look like years (e.g., 2010, X2010)
# -----------------------------
get_year_cols <- function(nm) {
  nm <- as.character(nm)
  is_year <- str_detect(nm, "^(X)?(19|20)\\d{2}$")
  nm[is_year]
}

# -----------------------------
# Helper: sanitize column names so dplyr/tibble never sees NA/blank names
# (Fixes: "Can't transform a data frame with `NA` or `\"\"` names.")
# -----------------------------
sanitize_names <- function(x) {
  x <- as.character(x)
  x <- str_trim(x)

  # Replace NA / empty with stable placeholders
  bad <- is.na(x) | x == ""
  if (any(bad)) {
    x[bad] <- paste0("col_", which(bad))
  }

  # Make syntactic + unique
  x <- make.names(x, unique = TRUE)

  # Light cleanup for readability
  x <- str_replace_all(x, "\\.+", "_")
  x <- str_replace_all(x, "^_|_$", "")
  x
}

# -----------------------------
# Helper: safely compute correlation matrix (returns NULL if insufficient numeric data)
# -----------------------------
safe_cor <- function(df_numeric, method = "pearson") {
  if (is.null(df_numeric) || ncol(df_numeric) < 2) return(NULL)

  df_num <- df_numeric %>%
    mutate(across(everything(), as.numeric)) %>%
    select(where(~ !all(is.na(.x))))

  if (ncol(df_num) < 2) return(NULL)

  tryCatch(
    stats::cor(df_num, use = "pairwise.complete.obs", method = method),
    error = function(e) NULL
  )
}
```

---

## 2) Load data (robust Excel handling)

This project uses an Excel export from DataWeb. Exported files can differ by:
- sheet names
- header row location
- presence of “metadata” rows above the year-by-year table

The loader below scans each sheet and finds the row that looks most like a year header (2010–2024 style).
It then reads the table starting at that header row.

```{r}
candidate_files <- c(
  "data/DataWeb-Query-Export_us_exports_total_2010_2024.xlsx",
  "data/us_exports.xlsx",
  "data/exports.xlsx",
  "data/dataweb.xlsx"
)

file_used <- first_existing(candidate_files)
file_used
```

```{r}
# If no file exists, stop early but keep the notebook readable.
if (is.na(file_used)) {
  stop("No exports Excel file was found in /data. Add the DataWeb export and re-render.")
}

sheets <- readxl::excel_sheets(file_used)

# Detect which row in a sheet is most likely the header row by counting "year-like" values
detect_header_row <- function(path, sheet, max_preview_rows = 40) {
  preview <- readxl::read_excel(
    path,
    sheet = sheet,
    col_names = FALSE,
    n_max = max_preview_rows
  )

  # Count how many year-like tokens appear in each row
  year_counts <- map_int(seq_len(nrow(preview)), function(r) {
    row_vals <- as.character(unlist(preview[r, ], use.names = FALSE))
    row_vals <- str_trim(row_vals)
    sum(str_detect(row_vals, "^(X)?(19|20)\\d{2}$"), na.rm = TRUE)
  })

  best_row <- which.max(year_counts)
  # If nothing looks like a header, return NA
  if (max(year_counts, na.rm = TRUE) == 0) return(NA_integer_)
  best_row
}

sheet_audit <- tibble(
  sheet = sheets,
  header_row = map_int(sheets, ~ detect_header_row(file_used, .x)),
  selected = header_row == max(header_row, na.rm = TRUE)
)

sheet_audit
```

```{r}
# Choose the sheet with the earliest (highest confidence) detected header row.
# (If multiple, pick the first.)
chosen_sheet <- sheet_audit %>%
  filter(!is.na(header_row)) %>%
  arrange(header_row) %>%
  slice(1) %>%
  pull(sheet)

chosen_sheet
```

```{r}
# Read the chosen sheet starting at the detected header row.
hdr_row <- sheet_audit %>% filter(sheet == chosen_sheet) %>% pull(header_row)

raw_block <- readxl::read_excel(
  file_used,
  sheet = chosen_sheet,
  skip = hdr_row - 1,
  col_names = FALSE
)

# First row of this block contains the headers (possibly with blanks)
header_vec <- as.character(unlist(raw_block[1, ], use.names = FALSE))
names(raw_block) <- sanitize_names(header_vec)

# Remove the header row from the data
raw_tbl <- raw_block[-1, ] %>% as_tibble()

glimpse(raw_tbl)
```

---

## 3) Clean + tidy to long format

Goal: build a consistent table with columns:

- **Entity** (destination / partner / category)
- **Year**
- **Exports** (numeric)

```{r}
# Heuristic: the "Entity" column is the first non-year column.
year_cols <- get_year_cols(names(raw_tbl))
non_year_cols <- setdiff(names(raw_tbl), year_cols)

entity_col <- non_year_cols[1]
entity_col
```

```{r}
# Build long table
exports_long <- raw_tbl %>%
  rename(Entity = all_of(entity_col)) %>%
  mutate(Entity = as.character(Entity)) %>%
  pivot_longer(
    cols = all_of(year_cols),
    names_to = "Year",
    values_to = "Exports"
  ) %>%
  mutate(
    Year = as.integer(str_remove(Year, "^X")),
    Exports = as.numeric(Exports)
  ) %>%
  filter(!is.na(Entity), Entity != "", !is.na(Year))

# Quick check
exports_long %>% slice_head(n = 10)
```

---

## 4) Trend view for top destinations

To keep multivariate visuals readable, focus on the **top N entities** by total exports across all years.

```{r}
top_n <- 10

top_entities <- exports_long %>%
  group_by(Entity) %>%
  summarize(total_exports = sum(Exports, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_exports)) %>%
  slice_head(n = top_n) %>%
  pull(Entity)

top_entities
```

```{r}
trend_df <- exports_long %>%
  filter(Entity %in% top_entities, !is.na(Exports)) %>%
  group_by(Entity, Year) %>%
  summarize(Exports = sum(Exports, na.rm = TRUE), .groups = "drop")

ggplot(trend_df, aes(x = Year, y = Exports, color = Entity)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Top export destinations over time",
    subtitle = paste0("Top ", top_n, " entities by total exports (all years combined)"),
    x = "Year",
    y = "Exports (as provided in DataWeb export)"
  ) +
  theme_minimal()
```

---

## 5) Build a “wide” numeric dataset for correlations

Wide format requirement:

- rows = **Year**
- columns = export series for each **Entity**
- values = **Exports**

```{r}
wide_numeric <- exports_long %>%
  filter(Entity %in% top_entities) %>%
  group_by(Entity, Year) %>%
  summarize(Exports = sum(Exports, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Entity, values_from = Exports) %>%
  arrange(Year)

wide_numeric %>% slice_head(n = 5)
```

---

## 6) Correlation matrix (multivariate)

```{r}
x_mat <- wide_numeric %>% select(-Year)
corr_mat <- safe_cor(x_mat, method = "pearson")

if (is.null(corr_mat)) {
  stop("Correlation matrix could not be computed (not enough numeric data).")
}

corr_mat[1:min(5, nrow(corr_mat)), 1:min(5, ncol(corr_mat))]
```

---

## 7) Visualize correlations

### 7A) Heatmap (ggplot)

```{r}
corr_df <- as.data.frame(corr_mat) %>%
  tibble::rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")

ggplot(corr_df, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", Correlation)), size = 3) +
  scale_fill_gradient2(limits = c(-1, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Correlation heatmap (Pearson)",
    subtitle = "Top export destinations — strength of co-movement across years",
    x = NULL,
    y = NULL
  )
```

### 7B) corrplot (only if installed)

```{r}
if (use_pkg("corrplot")) {
  corrplot::corrplot(
    corr_mat,
    method = "color",
    type = "upper",
    addCoef.col = "black",
    tl.cex = 0.8,
    number.cex = 0.7
  )
}
```

### 7C) Scatterplot matrix (optional; only if GGally installed)

```{r}
if (use_pkg("GGally")) {
  # ggpairs can be heavy. Keep it small and numeric.
  x_small <- x_mat
  if (ncol(x_small) >= 2) {
    GGally::ggpairs(x_small)
  }
}
```

---

## 8) Interpretation (STAR: Scope + Action)

### Scope

The dataset summarizes U.S. exports over time, grouped by destination/partner (depending on the DataWeb export configuration).
Multivariate correlation helps identify which destinations tend to increase/decrease together over the same years.

### Action

1. A time-series view identifies the most influential destinations (top N).
2. A wide numeric dataset is created to compute Pearson correlations.
3. The correlation matrix and heatmap reveal pairs (or clusters) of destinations with strong co-movement,
   which can support hypotheses about shared economic forces (global demand cycles, commodity effects, shipping constraints, etc.).

---

## 9) Economy & Labor tie-in (implemented)

A practical next step is to connect exports to a labor-market indicator.

This section uses the built-in `ggplot2::economics` dataset (monthly U.S. macro series) to compute an **annual unemployment rate average**,
then merges it with annual exports.

This produces:
- Correlation between total exports and unemployment
- A simple regression model for interpretability

```{r}
# 9A) Compute annual unemployment rate (from ggplot2 economics dataset)
labor_annual <- ggplot2::economics %>%
  mutate(
    Year = lubridate::year(date)
  ) %>%
  group_by(Year) %>%
  summarize(
    unemployment_avg = mean(unemploy, na.rm = TRUE),
    .groups = "drop"
  )

labor_annual %>% slice_head(n = 5)
```

```{r}
# 9B) Compute annual total exports (sum across all entities for each year)
exports_total_annual <- exports_long %>%
  group_by(Year) %>%
  summarize(
    exports_total = sum(Exports, na.rm = TRUE),
    .groups = "drop"
  )

exports_total_annual %>% slice_head(n = 5)
```

```{r}
# 9C) Merge and analyze
econ_labor_df <- exports_total_annual %>%
  inner_join(labor_annual, by = "Year") %>%
  arrange(Year)

econ_labor_df %>% slice_head(n = 10)
```

```{r}
# 9D) Visualization: exports vs unemployment (annual)
ggplot(econ_labor_df, aes(x = unemployment_avg, y = exports_total)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Exports vs labor conditions (annual)",
    subtitle = "Annual exports total vs annual average unemployment (from ggplot2::economics)",
    x = "Annual average unemployment (count, economics dataset)",
    y = "Annual total exports"
  ) +
  theme_minimal()
```

```{r}
# 9E) Correlation + regression
cor_val <- cor(econ_labor_df$exports_total, econ_labor_df$unemployment_avg, use = "complete.obs")
cor_val
```

```{r}
lm_fit <- lm(exports_total ~ unemployment_avg, data = econ_labor_df)
summary(lm_fit)
```

### Interpretation

- If the correlation is **negative**, higher unemployment years tend to align with lower export totals (consistent with weaker demand).
- If the correlation is **positive**, the time window may include structural shifts (e.g., exports rising while unemployment rises) and should be checked alongside other macro indicators (GDP, industrial production, inflation).

---

## 10) Forward momentum for the project

This week’s correlation structure provides a foundation for a predictive model.

1. **Expand the labor linkage** using additional annual indicators (labor-force participation, CPI, GDP, JOLTS).
2. Move from correlation to **multilinear regression**, using top destinations as predictors and a macro outcome as the response.
3. Validate assumptions (collinearity, residual diagnostics) and compare model variants (AIC / adjusted R²).
